Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-09-21 manuscript/manuscript.tex
Edit the D-delta search section
Removed a lot of text from this section. As I was reading through it, I
didn't find the explanation about the different sections of the grid
necessary. In fact, it made the text more complicated than it needed to
be. The justification that we need was only:

1. We want the lowest D and highest delta
2. Anything out of the gray line is ruled out
3. D_linear is the best D for the exponential case so we can fix it to
   that.

This is pretty much what that whole explanation was about. Since we
don't need to explain the regions anymore, we don't really need the
extr

Lets perform a similar comparison between the numerical model and
the analytical solution but now for a spherical shell with exponential
density.
On this scenario the problem is more complex due to the fact that the
density-based discretization will be certainly applied together with
the modified adaptive discretization algorithm.
This means that both the distance-size ratio $D$ and the delta ratio
$\delta$ must be simultaneously determined.

\subsubsection{$D$, $\delta$ space exploration}

From every combination of the $D$ and $\delta$ ratios we want to find
the one that produces a numerical error lower than the 0.1\% threshold
while keeping an effective computation, i.e. the ($D$, $\delta)$ pair
that produces the minimum number of discretizations and ensures an
acceptable numerical precision.
We can propose an heuristic method to perform this task: compute the
numerical error for every ($D$, $\delta$) points belonging to a grid on
the $D$, $\delta$ space and choose the pair that combines an
acceptable numerical error with an efficient computation.

Firstly, lets consider a decreasing exponential density function that
assumes the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's
outer and inner radii, respectively, that can be defined as follows:

\begin{equation}
    \rho(r') = A e^{-(r' - R)/b} + C,
\label{eq:density-exp}
\end{equation}
\noindent where
\begin{equation}
    A =
    (3300 \text{kg/m}^3 - 2670 \text{kg/m}^3)
    \left( e^{( R_2 - R_1 )/b} - 1 \right)^{-1},
\end{equation}
\begin{equation}
    C =
    2670 \text{kg/m}^3 - A,
\end{equation}

\noindent $R$ is the mean Earth radius and $b$ is a constant
that determines the variability of the function: a low value of $b$
increases the maximum slope of the density function.

Secondly, we propose a thick spherical shell with a thickness of 35km and
modelled by a mesh of $30^\circ \times 30^\circ$ tesseroids in order to
perform the exploration of the numerical error in the $D$, $\delta$
space.
We have assigned the density function proposed on equation
\ref{eq:density-exp} to our spherical shell model with a $b$ value of
1km, i.e. a highly varying density function.

Finally, we have computed the gravitational potential ($V$), the vertical
component of its gradient ($g_z$) and the $g_{zz}$ component of the
Marussi tensor on the big grid (see Table \ref{tab:grids}) by exploring
different values of $D$ and $\delta$.
These results were used to obtain the percentage difference between
them and their corresponding analytical solutions.

On Figure \ref{fig:grid-search} we show
the resulting percentage difference for each gravitational field.
The points inside the dotted lines are the ones that present a
numerical error lower than the 0.1\% threshold.
Our intention is to choose the one that produces the most efficient computation
and set it as the default values for $D$ and $\delta$.

\begin{figure}
\centering
\includegraphics[width=\linewidth]
    {figures/grid-search.pdf}
\caption{
    Numerical error exploration in the $D$, $\delta$ space.
    The percentage difference values were obtained from the comparison
    between the analytical solution and the numerical approximation of
    the gravitational fields ($V$, $g_z$ and $g_{zz}$) generated by a
    spherical shell with an exponential density as the one defined on
    equation \ref{eq:density-exp}.
    These comparison have been carried out with a spherical shell of
    35km of thickness, whose density function has a value of $b$ of 1km
    and the gravitational fields were computed on the big grid
    (defined in Table \ref{tab:grids}).
    The points inside the dashed line are the ones that present an
    error lower than 0.1\%.
    }
\label{fig:grid-search}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]
    {figures/D-delta-grid-search.pdf}
\caption{
    $D$, $\delta$ space divided in three regions (A, B and C) based on
    the definitions of $D_\text{linear}$ and $\delta_p$ (see text for more
    information).
    Light grey points are the non-eligible ($D$, $\delta$) pairs
    belonging to the region A (with $D < D_\text{linear}$).
    Green points have acceptable numerical error (lower than 0.1\%),
    while the red ones represent those whose error is greater than the
    threshold.
    Dark grey points represent those from which we don't have a priori
    information about their numerical error.}
\label{fig:D-delta-grid-search}
\end{figure}

Because the algorithm is thought to work with default $D$ and $\delta$
values that ensures these two conditions for every density function,
the default value for $D$ cannot be lower than the one determined for
the linear function, which will be called $D_\text{linear}$.
That's why we can only choose the ($D$, $\delta$) points on which
$D \geq D_\text{linear}$.
Taking this fact into account we can divide the $D$, $\delta$ space in
two subregions: one with $D < D_\text{linear}$ and other with $D \ge
D_\text{linear}$.
We can ignore points belonging to the former one (region A on Figure
\ref{fig:D-delta-grid-search}) and focus on the latter (regions B and C on
Figure \ref{fig:D-delta-grid-search}).

From the $D \ge D_\text{linear}$ subregion we can explore the numerical
error of points with $D = D_\text{linear}$ and different values of
$\delta$.
We will call $\delta_p$ to the maximum $\delta$ at which the numerical
error is lower than the 0.1\% threshold while keeping
$D = D_ \text{linear}$.
Greater $\delta$ values than $\delta_p$ will produce computations with
high error (red points in Figure \ref{fig:D-delta-grid-search}), while
lower $\delta$ values than $\delta_p$ will produce more discretizations
leading to non-efficient computation (green dots under the
$(D_\text{linear}, \delta_p)$ point on Figure \ref{fig:D-delta-grid-search}).

With the definition of $\delta_p$ in mind we can subdivide the
subregion in two: one with points with $\delta \le \delta_p$ (region B
on Figure \ref{fig:D-delta-grid-search}) and other with points with
$\delta > \delta_p$ (region C on Figure \ref{fig:D-delta-grid-search}).

Every point in the subregion with $\delta \le \delta_p$ (region B on
Figure \ref{fig:D-delta-grid-search}) will have a numerical error lower
than the 0.1\% threshold, but they will always produce more
discretizations than the ($D_\text{linear}$, $\delta_p$) point leading
to non-efficient computations.
It's worth remembering that a lower $\delta$ or a greter $D$ will
produce more discretizations, increasing the accuracy but also the
computation time.

On the other hand, we have no a priori information about the numerical
error of points with $\delta > \delta_p$ (region C on
Figure \ref{fig:D-delta-grid-search}).
Nevertheless, the numerical error exploration shown on
Figure \ref{fig:grid-search} doesn't present any
point in this region with error lower than the threshold.
Even though we are analysing a single case, it proves that no point in
region C is eligible as default ($D$, $\delta$) pair because it should
ensure numerical accuracy for every exponential density scenario.

In conclusion, the ($D$, $\delta$) point that minimises the number of
discretizations while keeping an acceptable numerical error under the
0.1\% threshold is the one whose ratio $D$ is equal to
$D_\text{linear}$ and its delta ratio is $\delta_p$.

@@

@@

The analytical solution of the gravitational potential generated by a
spherical shell with a density function given by equation
\ref{eq:density-exp} is

>>>>

For an exponential density function, the density-based discretization will be applied
before the adaptive discretization algorithm.
This means that optimal values for both the distance-size ratio $D$ and the delta ratio
$\delta$ must be determined.
We perform an error analysis similar to what was done for the linear density case.
However, we only consider the "Big Grid" scenario from Table \ref{tab:grids} for the
sake of brevity.

The spherical shell and tesseroid model will have an exponential density function given
by

\begin{equation}
    A =
    (3300 \text{kg/m}^3 - 2670 \text{kg/m}^3)
    \left( e^{( R - R_1 )/b} - 1 \right)^{-1},
\end{equation}

\begin{equation}
    C =
    2670 \text{kg/m}^3 - A,
\end{equation}

\noindent $R$ is the mean Earth radius, $R_1$ is the inner radius of the
shell (determined by the thickness), and $b$ is a constant that determines the
variability of the function: a low value of $b$ increases the maximum slope of the
density function.

\subsubsection{$D$-$\delta$ space exploration}

We aim to find a combination of the $D$ and $\delta$ that produces a numerical error
lower than the 0.1\% threshold while minimizing computation time.
We use a grid search method and compute the numerical error for every ($D$, $\delta$)
pair belonging to a grid on the $D$-$\delta$ space (Fig. \ref{fig:grid-search}).
Because this is a time consuming computation, we limit the analysis to a shell of 35 km
thickness and $b = 1\ \text{km}$ (Eq. \ref{eq:density-exp}), which yields a
sharp density variation within the shell.
For optimum algorithm performance, we search for the smallest possible value of $D$ and
the highest possible value of $\delta$.

We compute the relative difference between the numerical and analytical results for the
gravitational potential ($V$), the vertical component of its gradient ($g_z$), and the
$g_{zz}$ component of the Marussi tensor.
The results are shown in Fig. \ref{fig:grid-search}, in which points inside the dotted
lines are the ones that present a numerical error lower than the 0.1\% threshold.
Also shown in Fig. \ref{fig:grid-search} are the $D$ values determined for the linear
density function in the previous section ($D_\text{linear}$).
The values of $D_\text{linear}$ are within the 0.1\% threshold and are the optimum
values of $D$ for $V$ and $g_z$.
The results for $g_{zz}$ are in agreement with those obtained for the linear density
function in the case of a thick shell (Fig. \ref{fig:D-linear}).
Thus, we also choose the conservative value of $D=8$ as the optimum value for $g_{zz}$.
These results indicate that the values of $D_\text{linear}$ can be safely extrapolated
to the exponential case.

Having well controlled values for $D$, we are free to explore the integration error as a
function of $\delta$ in more detail and how it varies for different values of $b$ (Eq.
\ref{eq:density-exp}).

\begin{figure}
\centering
\includegraphics[width=\linewidth]
    {figures/grid-search.pdf}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison
    between the analytical solution and the numerical approximation of
    the gravitational fields ($V$, $g_z$ and $g_{zz}$) generated by a
    spherical shell with an exponential density function (Eq. \ref{eq:density-exp}).
    These comparisons were carried out on the "Big Grid" configuration (Table
    \ref{tab:grids}), with a spherical shell of
    35km of thickness, and density function with $b$ of 1km.
    The points inside the dashed line are the ones that present an
    error lower than 0.1\%.
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{equation}
    \rho(r') = A e^{-(r' - R)/b} + C,
\end{equation}

@@

The analytical solution of the gravitational potential generated by a
spherical shell with a density function is
